---
title: "Week 2: Mortality I, Exercise"
subtitle: "SOC6708 ADA"
author: "Monica Alexander and Hasani Rose"
format: pdf
execute: 
  warning: false
  message: false
  
---

# Installing packages and reading in data

```{r}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
```

For this example we are using mortality in Canada by cause, 2012-2016. Available from [StatCan](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310039201). 

```{r}
d1 <- read_csv(here("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/CAN_age_cod.csv"))
head(d1)
```


# Creating a life table

```{r}
dl <- d1 %>% 
  filter(year == 2016, cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, Mx)
head(dl)
 
```

We need to create columns:

- $n$
- $_na_x$
- $_nq_x$
- $_nd_x$
- $_nL_x$
- $T_x$
- $e_x$

```{r}
d_lt <- dl |> 
     mutate(n = case_when(
       age==0 ~ 1,
       age==1 ~ 4,
       TRUE ~ 5
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx,
       age==1 ~ 1.5,
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      qx = n * Mx / (1 + (n - ax)* Mx), # probability of death
      px = 1 - qx, # probability of survival
      lx = lag(cumprod(px), default = 1), # l(x+n) = lx*px
      dx = lx - lead(lx, default = 0), # ndx = lx - l(x+n)
      Lx = n * lead(lx, default = 0) + (ax* dx), # nLx = n*l(x+n) + nax*ndx
      Tx = rev(cumsum(rev(Lx))), # sum of Lx above age x
      ex = Tx / lx
      ) 

view(d_lt)
```

## What's the life expectancy at age 10?
72.715225

## Exercise

Calculate the lifespan disparity for Canada in 2016.
```{r}
d_lt |> 
  mutate(prod = dx*ex) |> 
  summarize(sum(prod))
# 12.45 years	
```
12.45 years	

# Now calculate life expectancy if all intentional injuries were deleted. Get the data we need:
```{r}
dls <- d1 %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]"|cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]") %>%  # Keeping only the year and two causes of death we need
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, cause, Mx) %>% # Keeping age, cause, and mortality rate
  mutate(cause = ifelse(cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]", "suicide", "total")) %>%  # Recode the cause labels to something short
  spread(cause, Mx) %>% # turning it from long to wide; now each age has both rates available in                          the same row
  rename(Mx_i = suicide,
         Mx = total)

head(dls)
```



## Exercise: Cause-deleted life expectancy

You need to create the same columns as above, but with the cause-deleted versions ($-i$). Do this by first creating the ratio $R_x^{-i} = \frac{M_x^{-i}}{M_x}$, use this to get $_nq_x^{-i}$, and the rest is the same. 

What's the cause-deleted life expectancy at age 10? What's the implied life lost due to suicide?

```{r}
dls <- dls |> 
  mutate(Rx_minusi = (Mx-Mx_i)/Mx) #total mortality - mortality due to the cause / total mortality
```

```{r}
d_lt2 <- dls |> 
     mutate(n = case_when(  
       age==0 ~ 1, 
       age==1 ~ 4, 
       TRUE ~ 5 
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx, 
       age==1 ~ 1.5, 
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      qx = n * Mx / (1 + (n - ax)* Mx), #probability of death
     qx_minusi = qx*Rx_minusi,
      px = 1 - qx,  #probability of survival
      lx = lag(cumprod(px), default = 1), #l(x+n) = lx*px
      dx = lx - lead(lx, default = 0), #number of deaths in time period
      Lx = n * lead(lx, default = 0) + (ax* dx), #Person Years Lived
      Tx = rev(cumsum(rev(Lx))), #Person Years Lived above age x, reverse cumulative sum reversed
      ex = Tx / lx, #Life Expectancy
     
     # CAUSE DELETED COLUMNS; removing suicide
     px_minusi = 1 - qx_minusi,  #probability of survival
      lx_minusi = lag(cumprod(px_minusi), default = 1), #l(x+n) = lx*px
      dx_minusi = lx_minusi - lead(lx_minusi, default = 0), #number of deaths in time period
      Lx_minusi = n * lead(lx_minusi, default = 0) + (ax* dx_minusi), #Person Years Lived
      Tx_minusi = rev(cumsum(rev(Lx_minusi))), #Person Years Lived above age x, reverse cumulative                                                 sum reversed
      ex_minusi = Tx_minusi / lx_minusi #Life Expectancy
      ) 

d_lt2 |>
  filter(age==10) |>
  select(age, ex, ex_minusi)

73.0282-72.71523 
0.31297 


```
If suicide were eliminated, a 10-year-old would expect to live about 73.0 more years instead of 72.7. The implied life lost due to suicide at age 10 is about 0.3 years.



END


