---
title: "Week 3: Mortality II"
subtitle: "SOC6708 ADA"
author: "Monica Alexander and Hasani Rose"
format: pdf
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(here)
library(readxl)
library(janitor)

```

# Decomposition

Let's read in WPP data from the first week week and calculate the age-specific mortality rates:

```{r}
# Population data
d_male <- read_xlsx(here("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/WPP2024_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx"), skip = 16)
d_male$sex <- "Male"
d_male <- d_male |> drop_na(Year)

d_female <- read_xlsx(here("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/WPP2024_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx"), skip = 16)
d_female$sex <- "Female"
d_female <- d_female |> drop_na(Year)

d <- rbind(d_male, d_female)
rm(d_male, d_female)

# Mortality data
d_male <- read_xlsx("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/WPP2024_MORT_F01_2_DEATHS_SINGLE_AGE_MALE.xlsx", skip = 16)
d_male$sex <- "Male"
d_male <- d_male |> drop_na(Year)

d_female <- read_xlsx("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/WPP2024_MORT_F01_3_DEATHS_SINGLE_AGE_FEMALE.xlsx", skip = 16)
d_female$sex <- "Female"
d_female <- d_female |> drop_na(Year)

dm <- rbind(d_male, d_female)
rm(d_male, d_female)


d <- d  |> 
  clean_names() |> 
  select(region_subregion_country_or_area, iso3_alpha_code, year, x0:sex) |>
  rename(region = region_subregion_country_or_area) |> 
  mutate(across(x0:x100, as.numeric))


dm <- dm  |> 
  clean_names() |> 
  select(region_subregion_country_or_area, iso3_alpha_code, year, x0:sex) |> 
  rename(region = region_subregion_country_or_area) |> 
  mutate(across(x0:x100, as.numeric))

d_long <- d  |>  
  pivot_longer(x0:x100, names_to = "age", values_to = "pop") |> # Making data long
  mutate(age = as.numeric(str_remove(age, "x")))

dm_long <- dm  |>  
  pivot_longer(x0:x100, names_to = "age", values_to = "deaths") |> 
  mutate(age = as.numeric(str_remove(age, "x")))

# join these two tibbles and calculate rates

asmr <- d_long |>  
  left_join(dm_long) |>  
  mutate(mx = deaths/pop) #age-specific mortality rate 

```

Do the decomposition of the difference between Kenya and Canada:

```{r}
asmr |> 
  #get data for Kenya
  filter(region == "Kenya",year==2023)  |> 
  select(sex, age, pop, mx) |> 
  rename(pop_kenya = pop, mx_kenya = mx) |> 
  #get data for Canada
  left_join(asmr |> 
              filter(region == "Canada",year==2023)  |> 
              select(sex, age, pop, mx) |> 
              rename(pop_can = pop, mx_can = mx) ) |> 
  # calculate population proportions
  mutate(prop_kenya = pop_kenya/sum(pop_kenya),
         prop_can = pop_can/sum(pop_can)) |> 
  mutate(rate_diff = mx_kenya - mx_can, # difference in mortality rates
         prop_diff = prop_kenya - prop_can) |> # different in the pop proportions
  mutate(ave_rate = (mx_kenya+mx_can)/2, # average mortality rate
         ave_prop = (prop_kenya+prop_can)/2) |> #average pop proportion
  mutate(age_contr = prop_diff*ave_rate, # age contribution
         rate_contr = rate_diff*ave_prop) |> # mortality contribution
  summarize(age_total_contr = sum(age_contr), # summing up over age
            rate_total_contr = sum(rate_contr)) |> 
  mutate(total_diff = age_total_contr+rate_total_contr)


```
age_total_contr = −0.0107; The negative sign means Kenya’s age structure lowers its crude death rate relative to Canada. Kenya has a much younger population.

rate_total_contr = +0.00998;  The positive sign means Kenya’s age-specific mortality risks are higher than Canada’s. Higher risks push Kenya’s crude death rate upward relative to Canada

total_diff = −0.000724; This is the net difference in crude death rates. Kenya’s crude death rate is only slightly lower than Canada’s, not because mortality is better, but because its youthful age structure masks higher mortality risks

Check that the difference is actually the difference between the two CDRs

```{r}
asmr |> 
  filter(region == "Kenya"|region=="Canada",year==2023) |> 
  group_by(region) |> 
  summarize(cdr = sum(mx*pop)/sum(pop)) |> # Computes the crude death rate for each country
  summarise(diff = cdr[region=="Kenya"] - cdr[region=="Canada"]) # The difference
```
It is. 

## Exercise

Decompose the difference in CDRs between USA and Japan in the year 2023. Is the majority of the difference due to age structure or mortality?

```{r}
asmr |> 
  #get data for United States
  filter(region == "United States of America",year==2023)  |> 
  select(sex, age, pop, mx) |> 
  rename(pop_usa = pop, mx_usa = mx) |> 
  #get data for Canada
  left_join(asmr |> 
              filter(region == "Japan",year==2023)  |> 
              select(sex, age, pop, mx) |> 
              rename(pop_jpn = pop, mx_jpn = mx) ) |> 
  # calculate population proportions
  mutate(prop_usa = pop_usa/sum(pop_usa),
         prop_jpn = pop_jpn/sum(pop_jpn)) |> 
  mutate(rate_diff = mx_usa - mx_jpn, # difference in mortality rates
         prop_diff = prop_usa - prop_jpn) |> # different in the pop proportions
  mutate(ave_rate = (mx_usa+mx_jpn)/2, # average mortality rate
         ave_prop = (prop_usa+prop_jpn)/2) |> #average pop proportion
  mutate(age_contr = prop_diff*ave_rate, # age contribution
         rate_contr = rate_diff*ave_prop) |> # mortality contribution
  summarize(age_total_contr = sum(age_contr), # summing up over age
            rate_total_contr = sum(rate_contr)) |> 
  mutate(total_diff = age_total_contr+rate_total_contr)
```
The majority of the difference is due to age structure



# Mortality models

Read in mortality rates for Ontario. These data come from the [Canadian Human Mortality Database](http://www.bdlc.umontreal.ca/chmd/prov/ont/ont.htm). 

```{r}
dm2 <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm2)
```

## Gompertz

Let's fit a Gompertz model to Male mortality rates in the year 1950 from age 40. What is the interpretation of the coefficient estimates?

```{r}
# clean up a bit
dm2 <- dm2 |> 
  clean_names() |> 
  mutate(age = as.numeric(age))

df_1950_40 <- dm2 |> 
  filter(year==1950, age>39) |> 
  select(age, male) 

df_1950_40 |> 
  ggplot(aes(age, male)) + 
  geom_point()+
  scale_y_log10()

# remove above 100
df_1950_40 <- df_1950_40 |> 
  filter(age<100)

summary(lm(log(male)~age, data = df_1950_40))


```
Each additional year of age increases mortality by 8.6%


### Exercise

Now fit a Gompertz model to male mortality rates form age 40 in every year. Plot the estimated alpha and beta coefficients in a scatter plot, color the points by year. Comment on what you observe. 

```{r}
dm_to_fit <- dm2 |>   # Preparing the data to be modeled
  filter(age>39, age<100) |> 
  select(year, age, male)

coefs <- tibble() # creates an empty table to store results

years <- unique(dm_to_fit$year) # extracts all distinct years in the data

for(i in 1:length(years)){ # This loop runs once per year
  this_df <- dm_to_fit |>
    filter(year == years[i], !is.na(male), male > 0) # Subset data to a single year
    
    ####SOMETHING
  
  this_mod <- lm(log(male)~age, data = this_df) # Fit the Gompertz model
  
  these_coefs <- tibble(alpha = exp(coef(this_mod)[1]), beta = coef(this_mod)[2])
  
  coefs <- bind_rows(coefs, these_coefs)
  
}
  

```

We can see a decline in baseline adult mortality (α) alongside a gradual increase in the rate of  aging (β). Improvements in survival at adult ages combined with increasing compression of mortality into older age.


## Lee-Carter


Let's get the Lee-carter model parameters for Ontario. First, get the matrix of age-specific rates:

```{r}
m_tx <- dm2 |> 
  filter(age < 101) |> 
  select(year, age, male) |> 
  pivot_wider(names_from = "age", values_from = "male") |> 
  select(-year) |> 
  as.matrix()

ages <- 0:100
years <- unique(dm2$year)
```

log and demean those rates:

```{r}
logm_tx <- log(m_tx)
logm_tx[is.infinite(logm_tx)] <- min(logm_tx[!is.infinite(logm_tx)])
ax <- apply(logm_tx, 2, mean) # the mean across all years by age
```

Do the SVD

```{r}
swept_logm_tx <- sweep(logm_tx, 2, ax)

svd_mx <- svd(swept_logm_tx)

names(svd_mx)
svd_mx$d

bx <- svd_mx$v[, 1]/sum(svd_mx$v[, 1]) # should have 100 entries, looking at age groups
kt <- svd_mx$d[1] * svd_mx$u[, 1] * sum(svd_mx$v[, 1]) 

```

Plots!

```{r}
# plot ax

lc_age_df <- tibble(age = ages, ax = ax, bx = bx)
lc_time_df <- tibble(year = years, kt = kt)

ggplot(lc_age_df, aes(age, ax)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("ax values for Ontario")

ggplot(lc_age_df, aes(age, bx)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("bx values for Ontario")

ggplot(lc_time_df, aes(year, kt)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("kt values for Ontario")

```

let's look at the fit for a couple of years

```{r}
data_and_res <- dm2 |>  # Build a dataset with the observed data and Lee–Carter parameters
  filter(age < 101) |> 
  mutate(log_mx = log(male)) |> 
  left_join(lc_age_df) |>  # Adding age specific parameters
  left_join(lc_time_df) |> 
  mutate(lc_fit = ax + bx*kt)

data_and_res |>  # Plot observed vs fitted for selected years
  filter(year %in% c(1921, 1951, 1981, 2011, 2021)) |> 
  ggplot(aes(age, log_mx)) + geom_point() + 
  facet_wrap(~year) + 
  geom_line(aes(age, lc_fit), color = "red") + 
  ggtitle("Data and fits for Ontario")

```

### Exercise

Repeat the lee-carter model fitting exercise but just use mortality rates from 1970. Does this change the estimated rates? Does it do a better or worse job, or does it depend on the year? 

## filter ages and look years

```{r}
# Restrict the data
dm2_1970 <- dm2 |> 
  filter(year >= 1970, age < 101)
# Build mortality matrix
m_tx_1970 <- dm2_1970 |> 
  select(year, age, male) |> 
  pivot_wider(names_from = age, values_from = male) |> 
  select(-year) |> 
  as.matrix()

logm_tx_1970 <- log(m_tx_1970)
logm_tx_1970[is.infinite(logm_tx_1970)] <- 
  min(logm_tx_1970[!is.infinite(logm_tx_1970)])

ax_1970 <- apply(logm_tx_1970, 2, mean)

swept_1970 <- sweep(logm_tx_1970, 2, ax_1970)

svd_1970 <- svd(swept_1970)

bx_1970 <- svd_1970$v[,1] / sum(svd_1970$v[,1])
kt_1970 <- svd_1970$d[1] * svd_1970$u[,1] * sum(svd_1970$v[,1])

# Ploting

lc_age_1970 <- tibble(age = 0:100, ax = ax_1970, bx = bx_1970)
lc_time_1970 <- tibble(year = unique(dm2_1970$year), kt = kt_1970)

dm2_1970 |> 
  mutate(log_mx = log(male)) |> 
  left_join(lc_age_1970) |> 
  left_join(lc_time_1970) |> 
  mutate(lc_fit = ax + bx*kt) |> 
  filter(year %in% c(1971, 1981, 1991, 2001, 2011, 2021)) |> 
  ggplot(aes(age, log_mx)) +
  geom_point() +
  geom_line(aes(y = lc_fit), color = "red") +
  facet_wrap(~year)

```

It slightly changes bx. It does a better job for recent years.

