---
title: "Week 9: Intro to Bayes"
subtitle: "SOC6708 ADA"
author: "Monica Alexander and Hasani Rose"
format: pdf
execute: 
  warning: false
  message: false
---

# Happiness

## Question 1

Consider the happiness example from the lecture, with 118 out of 129 women indicating they are happy. We are interested in estimating $\theta$, which is the (true) proportion of women who are happy. Calculate the MLE estimate $\hat{\theta}$ and 95% confidence interval. 

```{r}
y <- 118
n <- 129

# mean
y/n

```
The estimated proportion of women who report being happy is: 0.915 (or 91.5%)

```{r}
p <- y/n
se <- sqrt(p*(1-p)/n)

p - 1.96*se
p + 1.96*se

```
The 95% confidence interval is: [0.867, 0.963]

## Question 2

Assume a Beta(1,1) prior on $\theta$. Calculate the posterior mean for $\hat{\theta}$ and 95% credible interval. 

```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 1
b <- 1

p_bayes <- (y+a)/(n+a+b)

qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)


```
The posterior mean is 0.908 (or 90.8%)
After combining the prior information and the observed data, the estimated proportion of women who are happy is about 90.8%

The 95% credible interval is: [0.854, 0.951]
The prior is weak and the sample is fairly large; hence, probably why the Bayesian and frequentist answers are very similar

## Question 3

Now assume a Beta(10,10) prior on $\theta$. What is the interpretation of this prior? Are we assuming we know more, less or the same amount of information as the prior used in Question 2?

```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 10
b <- 10

p_bayes <- (y+a)/(n+a+b)
p_bayes


qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)



```
Posterior mean is: posterior mean = 0.859

95% credible interval = [0.799, 0.910]

With the Beta(10,10) prior, before seeing the data, we believe the true value of  θ is probably around 0.5, and values near 0 or 1 are less likely.

With Beta(1,1), we had no strong prior belief; all values between 0 and 1 were equally likely. But with Beta(10,10), we do have a prior belief, and it is stronger: we think θ is likely to be near 0.5. Hence, we have more prior information

The posterior estimate is pulled downward relative to the sample proportion and the posterior from Question 2

```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 10
b <- 100


p_bayes <- (y+a)/(n+a+b)
p_bayes

qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)

```
Here we have a prior that assumes, before seeing any data, that about 9% of women are happy

10 “successes” and 100 “failures”, which is a lot of prior information

This pulls down the mean a lot


```{r}
library(tidyverse)
dp <- read_csv("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/wpp_pop_clean.csv")
df <- read_csv("C:/Users/hasan/OneDrive/Desktop/SOC PhD Stuff 2025, 2026/ADA/WPP2024_Demographic_Indicators_Medium.csv")

dpf <- dp |> filter(region=="Canada", sex == "Female") |> 
  pivot_longer(x0:x100) |> 
  mutate(age = as.numeric(str_remove(name, "x")))

dpm <- dp |> filter(region=="Canada", sex == "Male") |> 
  pivot_longer(x0:x100) |> 
  mutate(age = as.numeric(str_remove(name, "x")))

breaks <- c(seq(0,100, by = 5), Inf)
labels <- seq(0,100, by = 5)

# Create the age groups
dpf$age_group <- cut(dpf$age, 
                  breaks = breaks, 
                  labels = labels, 
                  right = FALSE) 

dpf5 <- dpf |> 
  group_by(year, age_group) |> 
  summarize(pop = sum(value))

dpm$age_group <- cut(dpm$age, 
                  breaks = breaks, 
                  labels = labels, 
                  right = FALSE) 

dpm5 <- dpm |> 
  group_by(year, age_group) |> 
  summarize(pop = sum(value))

dp5_23 <- dpf5 |> filter(year==2023)

C <- dp5_23$pop[dp5_23$age_group==0] + dpm5$pop[dpm5$age_group==0&dpm5$year==2023] # Children

W <- dp5_23$pop[dp5_23$age_group==15]+dp5_23$pop[dp5_23$age_group==20]+dp5_23$pop[dp5_23$age_group==25]+dp5_23$pop[dp5_23$age_group==30]+dp5_23$pop[dp5_23$age_group==35]+dp5_23$pop[dp5_23$age_group==40]+dp5_23$pop[dp5_23$age_group==45] # Women

p <- 1/7 # Assumes fertility is spread evenly across the 7 reproductive age groups
s <- 1 #  Assumes no child mortality

iTFR <- 1/s*1/p*C/W
iTFR
### Indirect TFR estimate IS 1.48

x_multiplier <- (10.65-12.55*(dp5_23$pop[dp5_23$age_group==25]+dp5_23$pop[dp5_23$age_group==30])/W) ## this formula adjusts based on the proportion of women in age groups 25–29, and 30–34
xTFR <- x_multiplier*C/W
xTFR


# the actual TFR is
df |> filter(Location=="Canada") |> select(Time, TFR)
df |> filter(Location=="Canada", Time == 2023) |> select(Time, TFR)
```
The simple iTFR estimate is 1.48

xTFR is 1.448, slightly lower than the simple iTFR.

The actual TFR for Canada in 2023 is 1.3494. Hence, both indirect estimates overestimated, though the xTFR was slightly closer



